Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex

@@

\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
  Software Product Sustainability Model~\cite{Calero_WSSSPE}

\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}

\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}

\item Daniel S. Katz, David Proctor. A Framework for Discussing
  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}

\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}

\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}

@@

\subsection{What are communities?}

The workshop did not directly answer the question ``What are communities?'' but
instead a number of different answers were indirectly presented, through the
depiction of individuals and stakeholders in different aspects of the
scientific software lifecycle.  In broad strokes, however, scientific software
communities were generally accepted as consisting of individuals, often but not
always composed of scientist practitioners, that were working with some degree
of coordination toward a common goal enabled by software.

The discussion of development-focused communities centered around describing
methods of interaction between individuals and the scientific software.  The
first type of interaction was the development of a specific piece of software,
the second was a particular domain or discipline, and the final primary
type of interaction was around the development of applications built on a
particular piece of software that was perhaps developed by another group.  As
an example, in \cite{Hanwell_WSSSPE}, the community described is comprised of
both the for-profit company Kitware and the users and contributors to their
software packages such as VTK.  This structure, of the centralized development
of core infrastructure around which communities of individuals applying that
infrastructure and developing applications utilizing it, was similarly
reflected in \cite{Terrel_WSSSPE}, where the core scientific python ecosystem
is supported by a non-profit entity that fosters community investment in that
ecosystem.  In many ways, these two organizations (Kitware and NumFOCUS)
attempt to cross domain boundaries and provide support for both the
infrastructure and application sides of community building.

\subsection{Challenges of community}
include metrics for community success (e.g., more external
contributors than internal).
\subsection{Admirable scientific software communities}
\begin{itemize}
\item example communities
\item their (quick) origin stories.
\end{itemize}
\subsection{Resources for learning about software communities}
\begin{itemize}
\item academic fields studying communities (and software communities)
\item courses on online communities
\item books
\item need for software carpentry module on organizing communities?
\end{itemize}

>>>>

% note that paper summaries are available in an earlier commit 
% in James Howison's fork:
% https://github.com/jameshowison/WSSSPE/blob/587705cb6f12f46951e822fdbf5eb72b104fa397/WSSSPE1_report/WSSSPE1_summary_report.tex

@@

Frameworks provide a scaffold for thinking; they help us understand which elements are important and how they relate. They also help us break down a problem into chunks that can be pursued by different groups or perhaps disciplines.

A group of papers submitted to the workshop provide frameworks with these intentions. The frameworks have substantial overlap but with different emphases and extents.  Accordingly we present them starting with the area of most overlap and proceed to show the different emphases and the range of factors included in the frameworks.

\todo{seems like definitions will be mentioned elsewhere in the paper?}
Each of the framework papers included a definition of sustainability, with many overlaps.  In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.

One area in which there was not complete overlap was whether the word (and thus the effort called for by the WSSSPE workshop) involved environmental sustainability. Of course the word sustainability has strong connotations from consideration of environmental issues, evoking some mention of the areas in which software interacts with overall environmental resource usage, particularly energy efficiency.  The two papers in this area which mentioned this did so without integrating that analysis into the question of software being around long-term, suggesting that questions of environmental impact of scientific software is a conceptually distinct area of inquiry.

One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality. This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.  Thus Calero and colleagues propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perduability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability. This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al. argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact. Lenhardt et al compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered.) In addition their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.  In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability. The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project). For example Pierce et al focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance. In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''. They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.  Katz and Proctor also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance~\cite{Katz_WSSSPE}.  They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).

Frameworks also addressed themselves to the context in which software projects exist (thus moving upwards in abstraction, from the artifact itself, to the organization of its production and, in these frameworks, the shape of the space which an artifact or project exists in). These frameworks take the form of contingency theories in that they outline a different set of challenges and argue that different project organizations (and presumably artifact attributes) are necessary to persist longterm in spaces with particular characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking of this idea of space in terms of three axes: temporal (long or short term needs), spatial (local or global use) and purpose (specific to general).  They propose that different project organizations will be needed in different locations and argue that we should concentrate research to understand those connections. Weber et al~\cite{Weber_WSSSPE} describe their spaces by analogy with natural ecosystems as ``niches'' which sustain particular pieces of software. They define ``a software niche as the set of technical requirements, organizational conditions and cultural mores that support its maintenance and use over time.'' They call for better understanding and modeling of niches (as well as further exploration of the usefulness of ecosystem metaphors).

@@

\subsection{What are communities?}
\subsection{Challenges of community}
include metrics for community success (e.g., more external
contributors than internal).
\subsection{Admirable scientific software communities}
\begin{itemize}
\item example communities
\item their (quick) origin stories.
\end{itemize}
\subsection{Resources for learning about software communities}
\begin{itemize}
\item academic fields studying communities (and software communities)
\item courses on online communities
\item books
\item need for software carpentry module on organizing communities?
\end{itemize}
